---
title: "BikeWise: How to Avoid the Rain"
author: "SebKrbs"
date: "May 31st, 2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BikeWise: Rain-Aware Cycling Route Planning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# This chunk runs silently and sets default options for all code chunks below
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## What is BikeWise?

BikeWise helps you decide when to leave for a bike ride by checking the weather
*along your route*, not just at your front door. If it's going to rain at the
halfway point in 20 minutes and you'll be there in 22, you probably want to know
that before you leave.

The app lets you save your common locations (home, work, etc.) and then tells you
either "go now" or "leave at X" based on a route-level rain forecast.

You can try it here without installing anything:
**[https://sebkrbs.shinyapps.io/BikeWise/](https://sebkrbs.shinyapps.io/BikeWise/)**

*(screenshot of the app to be added here)*

---

## How it works — overview

The package is built in five layers:

1. **User accounts** — usernames and hashed passwords stored in Google Sheets
2. **Location management** — save addresses by label, geocoded automatically
3. **Routing** — OSRM API gives a bike-friendly path and timing
4. **Rain checking** — Open-Meteo forecast checked at each point along the route
5. **Shiny app** — ties everything together in a mobile-style interface

---

## Installation

```{r install, eval = FALSE}
remotes::install_github("Programming-The-Next-Step-2026/BikeWise")
```

To use the package against your own Google Sheet you also need to authenticate
and set a sheet ID (see README for details). The hosted app above handles this
automatically.

---

## Main functions

The two most important functions for users are `bikeroute()` and `raintracker()`.

### bikeroute()

Takes two coordinate pairs and returns a cycling route from the OSRM API,
including a `timed_coords` data frame that records your estimated position every
2 minutes.

```{r bikeroute, eval = FALSE}
# eval = FALSE because this calls an external API — we don't want it running
# every time the vignette is built
route <- bikeroute(
  from_lat = 52.3731, from_lon = 4.8922,
  to_lat   = 52.3579, to_lon   = 4.8686
)

route$duration_min
route$distance_km
head(route$timed_coords)
```

### raintracker()

Takes the `timed_coords` output and checks the rain forecast at each checkpoint
at the time you'd actually be there. Returns a suggested departure time.

```{r raintracker, eval = FALSE}
result <- raintracker(
  timed_df   = route$timed_coords,
  start_time = Sys.time(),
  threshold  = "moderate"
)

result$safe_to_go
result$suggested_departure
result$end_of_route_note
```

The `threshold` parameter sets how much rain you're willing to cycle in:
`"none"`, `"light"`, `"moderate"`, or `"heavy"`.

### Supporting functions

- `authenticate_user(username, password)` — log in or create an account
- `save_location(user, label, address)` — geocode and save a location
- `get_locations(user)` — retrieve saved locations as a data frame
- `rain_tolerance(username)` / `rain_tolerance(username, tolerance)` — get or set rain tolerance

---

## Data sources

- **[Open-Meteo](https://open-meteo.com)** — precipitation forecasts, free, no API key
- **[OSRM](https://project-osrm.org)** — bike routing, free, no API key
- **[Nominatim](https://nominatim.openstreetmap.org)** — geocoding, free, no API key
